---
title: "Map analysis"
author: "Wesley Rancher"
date: "2025-05-08"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r}
library(terra)
library(dplyr)
library(ggplot2)
library(patchwork)
library(purrr)
library(viridis)
```

Files

```{r}
mapdir <- "/Users/wancher/Documents/thesis/data/MapAnalysis/"
#mapdir <- "F:/jetstream/data/Maps/"
#landis reference files
#ref <- rast(paste0(mapdir, "AK_ClimateMap_10_Regions.tif"))
#bounds <- vect(paste0(mapdir, "../map-files/FullLandscapeV3_082722.shp"))
ref <- rast("/Users/wancher/Documents/thesis/data/MapAnalysis/AK_ClimateMap_10_Regions.tif")
#bounds <- vect("F:/FullLandscapeV3_082722.shp")

#projection
albers <- "EPSG:3338"

#eric levenson's pond and lake dataset to mask
# pond_files <- list.files(paste0(mapdir, "Alaska_Lake_Pond_Occurrence_2399/data/"), pattern=".tif", full.names=TRUE)
# pond_collection <- sprc(lapply(pond_files[c(5,6)], rast))
# pond_m <- mosaic(pond_collection)
# crs(pond_m) <- albers
# ext(pond_m) <- ext(bounds)
# ponds_r <- mask(pond_m, bounds)
```

Define functions to manipulate raster data and convert it to dataframes

```{r setup, include=TRUE}
crs(ref) <- albers
ref_cleaned <- mask(ref, bounds)


reclass_matrix_land <- matrix(
  c(0.1, 400, 1,         
    400, 700, 2,     
    700, 1000, 3,      
    1000, 1500, 4,
    1500, 2500, 5,
    2500, 4500, 6
  ), ncol = 3, byrow = TRUE)

reclass_matrix_rs <- matrix(
  c(0.1, 50, 1,         
    50, 150, 2,     
    150, 300, 3,      
    300, 450, 4,
    450, 700, 5
  ), ncol = 3, byrow = TRUE)

project_landis <- function(file){
  r <- rast(file)
  crs(r) <- albers
  ext(r) <- ext(ref_cleaned)
  r <- flip(r, direction="vertical")
  r_clipped <- mask(r, bounds)
  
  #remove inactive cells and water bodies
  cleanr <- mask(r_clipped, ref_cleaned, maskvalues=0)
  classified_r <- classify(cleanr*0.47, reclass_matrix_land)
  #cleanr <- mask(cleanr, ponds_r)#lakes and ponds
  
  df <- as.data.frame(cleanr, xy = TRUE)
  df$Model <- "LANDIS-II"
  df$ID <- basename(file)
  print(dim(df))
  return(cleanr)
}
project_rs <- function(file){
  r <- rast(file)
  #30m to 200m :(
  resampled_r <- resample(r, ref_cleaned, method = "bilinear")
  #resampled_r <- aggregate(r, fact = 200 / res(r)[1], fun = mean)
  cleanr <- mask(resampled_r, ref_cleaned, maskvalues=0)
  classified_r <- classify(cleanr, reclass_matrix_rs)
  
  #dataframe
  df <- as.data.frame(resampled_r, xy = TRUE)
  df$Model <- "RF-RS"
  df$ID <- basename(file)
  print(dim(df))
  varnames(cleanr) <- basename(file)
  return(cleanr)
}

```

Apply functions to project and difference

```{r}
#landis identifiers
reps <- c(4,5,6)
yrs <- c(0:4,20:24)
sp <- c("BlackSpruce", "WhiteSpruce", "QuakingAspen", "PaperBirch")
combs <- expand.grid(rep = reps, sp = sp, yr = yrs, stringsAsFactors = FALSE)

landisdir <- "F:/Rancher_Sims_Full_Landscape/FinalSimulationsThesis/"

#files
rs_files <- list.files(mapdir, 
                       pattern = "MOSAIC_.*.tif", 
                       full.names = TRUE)

landis_files <- with(combs, sprintf("%sinterior_hist_ncar_rep%d/biomass/%s-%d.img", 
                                    landisdir, rep, sp, yr))

#apply the functions
rs_rasters <- lapply(rs_files, project_rs)
landis_rasters <- lapply(landis_files, project_landis)
```

Visualizing relative dominance change

```{r}
# Helper function to summarize raster
summarize_raster <- function(r, species, model, timepoint) {
  tibble(
    Model = model,
    Species = species,
    Year = timepoint,
    TotalC = global(r, sum, na.rm = TRUE)[[1]]
  )
}
```

Adding in some line by line median compositing

```{r}
#filter by var name and select
raster_names <- sapply(rs_rasters, function(x) varnames(x))
bs_rs_rasters <- grep("Black_spruce", raster_names)
ab_rs_rasters <- grep("Resin_birch", raster_names)
qa_rs_rasters <- grep("Quaking_aspen", raster_names)
ws_rs_rasters <- grep("White_spruce", raster_names)

#rasterize
bs_rs_stack <- rast(rs_rasters[bs_rs_rasters])
ab_rs_stack <- rast(rs_rasters[ab_rs_rasters])
qa_rs_stack <- rast(rs_rasters[qa_rs_rasters])
ws_rs_stack <- rast(rs_rasters[ws_rs_rasters])

#calculate medians
bs_rs_median_t1 <- app(bs_rs_stack[[1:5]], median, na.rm = TRUE)
bs_rs_median_t2 <- app(bs_rs_stack[[6:10]], median, na.rm = TRUE)
ab_rs_median_t1 <- app(ab_rs_stack[[1:5]], median, na.rm = TRUE)
ab_rs_median_t2 <- app(ab_rs_stack[[6:10]], median, na.rm = TRUE)
qa_rs_median_t1 <- app(qa_rs_stack[[1:5]], median, na.rm = TRUE)
qa_rs_median_t2 <- app(qa_rs_stack[[6:10]], median, na.rm = TRUE)
ws_rs_median_t1 <- app(ws_rs_stack[[1:5]], median, na.rm = TRUE)
ws_rs_median_t2 <- app(ws_rs_stack[[6:10]], median, na.rm = TRUE)


rs_summary <- bind_rows(
  summarize_raster(bs_rs_median_t1, "Black Spruce", "RS", "T1"),
  summarize_raster(bs_rs_median_t2, "Black Spruce", "RS", "T2"),
  summarize_raster(ab_rs_median_t1, "Alaskan Birch", "RS", "T1"),
  summarize_raster(ab_rs_median_t2, "Alaskan Birch", "RS", "T2"),
  summarize_raster(qa_rs_median_t1, "Trembling Aspen", "RS", "T1"),
  summarize_raster(qa_rs_median_t2, "Trembling Aspen", "RS", "T2"),
  summarize_raster(ws_rs_median_t1, "White Spruce", "RS", "T1"),
  summarize_raster(ws_rs_median_t2, "White Spruce", "RS", "T2")
)



#difference
bs_rs_median_diff <- bs_rs_median_t2 - bs_rs_median_t1
ab_rs_median_diff <- ab_rs_median_t2 - ab_rs_median_t1
qa_rs_median_diff <- qa_rs_median_t2 - qa_rs_median_t1
ws_rs_median_diff <- ws_rs_median_t2 - ws_rs_median_t1

#writeRaster(bs_rs_median_diff, paste0(mapdir, "Median_BlackSpruce_RS_Difference.tif"))
#writeRaster(ab_rs_median_diff, paste0(mapdir, "Median_AlaskanBirch_RS_Difference.tif"))
#writeRaster(qa_rs_median_diff, paste0(mapdir, "Median_QuakingAspen_RS_Difference.tif"))
#writeRaster(ws_rs_median_diff, paste0(mapdir, "Median_WhiteSpruce_RS_Difference.tif"))
```

Replicate averages and median composites for landis maps

```{r}
#filter by var name and select
landis_raster_names <- sapply(landis_rasters, function(x) varnames(x))
t1_years <- 0:4
t2_years <- 20:24

#empties for results
median_diff <- list()

for (s in sp) {
  # --- Step 1: average across replicates per year ---
  yearly_means <- list()
  for (yr in c(t1_years, t2_years)) {
    pattern <- paste0("^", s, "-", yr, "$")
    indices <- grep(pattern, landis_raster_names)
    
    if (length(indices) == 0) next 
    
    rasters_for_year <- rast(landis_rasters[indices])
    yearly_mean <- app(rasters_for_year, mean, na.rm = TRUE)
    yearly_means[[as.character(yr)]] <- yearly_mean
  }
  # --- Step 2: Stack yearly means for t1 and t2 ---
  t1_stack <- rast(yearly_means[as.character(t1_years)])
  t2_stack <- rast(yearly_means[as.character(t2_years)])

  # --- Step 3: Median across years ---
  t1_median <- app(t1_stack, median, na.rm = TRUE)
  t2_median <- app(t2_stack, median, na.rm = TRUE)

  # --- Step 4: Difference (T2 − T1) ---
  median_diff[[s]] <- t2_median - t1_median
}

#show
plot(median_diff[["BlackSpruce"]])

# LANDIS summaries
landis_summary <- bind_rows(
  summarize_raster(median_diff[["BlackSpruce"]] + bs_rs_median_t1, "Black Spruce", "LANDIS", "T2"),
  summarize_raster(bs_rs_median_t1, "Black Spruce", "LANDIS", "T1"),
  summarize_raster(median_diff[["PaperBirch"]] + ab_rs_median_t1, "Alaskan Birch", "LANDIS", "T2"),
  summarize_raster(ab_rs_median_t1, "Alaskan Birch", "LANDIS", "T1"),
  summarize_raster(median_diff[["QuakingAspen"]] + qa_rs_median_t1, "Trembling Aspen", "LANDIS", "T2"),
  summarize_raster(qa_rs_median_t1, "Trembling Aspen", "LANDIS", "T1"),
  summarize_raster(median_diff[["WhiteSpruce"]] + ws_rs_median_t1, "White Spruce", "LANDIS", "T2"),
  summarize_raster(ws_rs_median_t1, "White Spruce", "LANDIS", "T1")
)

#writeRaster(median_diff[["BlackSpruce"]], paste0(mapdir, "Median_BlackSpruce_LANDIS_Difference.tif"))
#writeRaster(median_diff[["PaperBirch"]], paste0(mapdir, "Median_AlaskanBirch_LANDIS_Difference.tif"))
#writeRaster(median_diff[["QuakingAspen"]], paste0(mapdir, "Median_QuakingAspen_LANDIS_Difference.tif"))
#writeRaster(median_diff[["WhiteSpruce"]], paste0(mapdir, "Median_WhiteSpruce_LANDIS_Difference.tif"))
```

Convert to list of difference rasters, normalize, and convert to df

```{r}
# diff_list <- list(
#   "Black Spruce_RS" = rs_rasters[[2]] - rs_rasters[[1]],
#   "Quaking Aspen_RS" = rs_rasters[[4]] - rs_rasters[[3]],
#   "Alaskan Birch_RS" = rs_rasters[[6]] - rs_rasters[[5]],
#   "White Spruce_RS" = rs_rasters[[8]] - rs_rasters[[7]],
#   "Black Spruce_LANDIS" = landis_rasters[[2]] - landis_rasters[[1]],
#   "Alaskan Birch_LANDIS" = landis_rasters[[4]] - landis_rasters[[3]],
#   "Quaking Aspen_LANDIS" = landis_rasters[[6]] - landis_rasters[[4]],
#   "White Spruce_LANDIS" = landis_rasters[[8]] - landis_rasters[[7]]
# )
# 
# full_list <- list(
#   "Black Spruce_RS 2000" = rs_rasters[[1]],
#   "Black Spruce_RS 2024" = rs_rasters[[2]],
#   "Quaking Aspen_RS 2000" = rs_rasters[[3]],
#   "Quaking Aspen_RS 2024" = rs_rasters[[4]],
#   "Alaskan Birch_RS 2000" = rs_rasters[[5]],
#   "Alaskan Birch_RS 2024" = rs_rasters[[6]],
#   "White Spruce_RS 2000" = rs_rasters[[7]],
#   "White Spruce_RS 2024" = rs_rasters[[8]],
#   "Black Spruce_LANDIS 2000" = landis_rasters[[1]],
#   "Black Spruce_LANDIS 2024" = landis_rasters[[2]],
#   "Quaking Aspen_LANDIS 2000" = landis_rasters[[3]],
#   "Quaking Aspen_LANDIS 2024" = landis_rasters[[4]],
#   "Alaskan Birch_LANDIS 2000" = landis_rasters[[5]],
#   "Alaskan Birch_LANDIS 2024" = landis_rasters[[6]],
#   "White Spruce_LANDIS 2000" = landis_rasters[[7]],
#   "White Spruce_LANDIS 2024" = landis_rasters[[8]]
# )

#read in processed median rasters
median_list <- list.files(paste0(mapdir, "Medians/"), full.names = TRUE)
median_rasters <- lapply(median_list,rast)
names(median_rasters) <- basename(median_list)

#grab inactive cells from climate regions map
bounds <- vect(paste0(mapdir, "../map-files/FullLandscapeV3_082722.shp"))
climregions <- rast(paste0(mapdir, "AK_ClimateMap_10_Regions.tif"))
crs(climregions) <- albers
clipped_clim <- mask(climregions, bounds)
clipped_clim[clipped_clim != 0] <- NA
zero_df <- as.data.frame(clipped_clim,xy=TRUE)

# convert to data frame
raster_df <- bind_rows(lapply(names(median_rasters), function(name) {
  df <- as.data.frame(median_rasters[[name]], xy = TRUE)
  df$Species <- strsplit(name, "_")[[1]][2]
  df$Model <- strsplit(name, "_")[[1]][3]
  #df$Year <- sub(".* (\\d{4})", "\\1", name)
  colnames(df)[3] <- "Value"
  return(df)
}))

#plotting
raster_df$Class <- factor(raster_df$Value)


#fill_scale <- scale_fill_distiller(palette="YlGnBu", direction=1,na.value="black")
fill_scale <- scale_fill_distiller(
  palette = "RdYlGn",  # Red → Yellow → Green
  direction = 1,       # Use -1 if you want Green → Red
  na.value = "black"
)

# Base theme
base_theme <- theme_minimal(base_family = "Times New Roman") +
  theme(
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    strip.text = element_text(color = "black", size = 13)
  )

# species name to lowercase
library(stringr)
library(sf)
bounds_sf <- st_read("/Users/wancher/Documents/thesis/data/map-files/FullLandscapeV3_082722.shp")

#rename quaking aspen
raster_df_processed_clean<- raster_df %>%
  mutate(Species = ifelse(Species == "AlaskanBirch", "Alaskan birch", Species),
         Species = ifelse(Species == "QuakingAspen", "Trembling aspen", Species),
         Species = ifelse(Species == "BlackSpruce", "Black spruce", Species),
         Species = ifelse(Species == "WhiteSpruce", "White spruce", Species))%>%
  mutate(Species = factor(Species, levels = c("Black spruce", 
                                              "White spruce", 
                                              "Alaskan birch", 
                                              "Trembling aspen")))
#RS plot
rs_df <- raster_df_processed_clean%>% filter(Model == "RS")
rs_plots <- list()
for (spp in species){
  rs_plot <- ggplot() +
    geom_raster(data = rs_df%>%filter(Species==spp), aes(x = x, y = y, fill = Value)) +
    geom_raster(data = zero_df, aes(x = x, y = y), fill = "black") +
    geom_sf(data = bounds_sf, fill = NA, color = "black", linewidth = 0.3) +
    fill_scale +
    coord_sf(expand = FALSE) +
    facet_grid(Species~.) +
    #labs(fill = "Δ C\n(g m-2)")+
    base_theme +
    theme(legend.position = "none")  
  rs_plots[[spp]] <- rs_plot
}


# LANDIS plot (has the legend)
landis_plots <- list()
landis_df <- raster_df_processed_clean %>% filter(Model == "LANDIS")
for (spp in species){
  landis_plot <- ggplot() +
    geom_raster(data=landis_df%>%filter(Species==spp), aes(x = x, y = y, fill = Value)) +
    geom_raster(data = zero_df, aes(x = x, y = y), fill = "black")+
    geom_sf(data = bounds_sf, fill = NA, color = "black", linewidth = 0.3) +
    fill_scale +
    coord_sf(expand = FALSE)+
    facet_grid(Species~.) +
    #labs(fill = "Δ C\n(g m-2)")+
    base_theme +
    theme(legend.position = "none")
  landis_plots[[spp]] <- landis_plot
}
# Combine plots
combined <- (rs_plot|landis_plot)
  #plot_layout(widths = c(1, 1), guides = "collect")

combined

rs_plots_reduced <- reduce(rs_plots, `/`)
landis_plots_reduced <- reduce(landis_plots, `/`)
combined_plot <- rs_plots_reduced | landis_plots_reduced
combined_plot
```

```{r}
png("figures/Results_Carbon_Maps_LANDIS.png", width = 5, height = 11, units = "in", res = 300)
print(landis_plots_reduced)
dev.off()
```

Hana says I need to visualize the change since there is a lot of green

```{r}
#libraries
library(tidyverse)
library(sf)
library(patchwork)
library(RColorBrewer)
library(terra)
library(dplyr)
library(ggplot2)
library(patchwork)
library(purrr)
library(viridis)

#raster directory
mapdir <- "/Users/wancher/Documents/thesis/data/MapAnalysis/"

#read in processed median rasters
median_list <- list.files(paste0(mapdir, "Medians/"), full.names = TRUE)
median_rasters <- lapply(median_list,rast)
names(median_rasters) <- basename(median_list)

# Read bounds
bounds_sf <- st_read("/Users/wancher/Documents/thesis/data/map-files/FullLandscapeV3_082722.shp")

# Convert median raster list to long data frame
raster_df <- bind_rows(lapply(names(median_rasters), function(name) {
  df <- as.data.frame(median_rasters[[name]], xy = TRUE)
  df$Species <- strsplit(name, "_")[[1]][2]
  df$Model <- strsplit(name, "_")[[1]][3]
  colnames(df)[3] <- "Value"
  return(df)
}))

# Classify and clean species names
raster_df_processed_clean <- raster_df %>%
  mutate(
    Species = case_when(
      Species == "AlaskanBirch" ~ "Alaskan birch",
      Species == "QuakingAspen" ~ "Trembling aspen",
      Species == "BlackSpruce" ~ "Black spruce",
      Species == "WhiteSpruce" ~ "White spruce",
      TRUE ~ Species
    ),
    Species = factor(Species, levels = c("Black spruce", "White spruce", "Alaskan birch", "Trembling aspen"))
  )
change_levels <- c("< -100", "-100 to -50", "-50 to -10", "-10 to -1",
                   "-1 to 1", "1 to 10", "10 to 50", "50 to 100", "> 100")
raster_df_processed_clean <- raster_df_processed_clean %>%
  mutate(ChangeClass = cut(
    Value,
    breaks = c(-Inf, -100, -50, -10, -1, 1, 10, 50, 100, Inf),
    labels = change_levels,
    include.lowest = TRUE
  ))
# All combinations of Species and ChangeClass
all_combinations <- expand.grid(
  Species = levels(raster_df_processed_clean$Species),
  ChangeClass = factor(change_levels, levels = change_levels)
)

#rs change summary
change_summary <- raster_df_processed_clean %>%
  filter(Model == "RS") %>%
  mutate(ChangeClass = factor(ChangeClass, levels = change_levels)) %>%
  group_by(Species, ChangeClass) %>%
  summarise(count = n(), .groups = "drop") %>%
  complete(Species, ChangeClass = factor(change_levels, levels = change_levels), fill = list(count = 0)) %>%
  group_by(Species) %>%
  mutate(
    percent = 100 * count / sum(count),
    ChangeClass = factor(ChangeClass, levels = change_levels)
  )
#landis 
change_summary_2 <- raster_df_processed_clean %>%
  filter(Model == "LANDIS") %>%
  mutate(ChangeClass = factor(ChangeClass, levels = change_levels)) %>%
  group_by(Species, ChangeClass) %>%
  summarise(count = n(), .groups = "drop") %>%
  complete(Species, ChangeClass = factor(change_levels, levels = change_levels), fill = list(count = 0)) %>%
  group_by(Species) %>%
  mutate(
    percent = 100 * count / sum(count),
    ChangeClass = factor(ChangeClass, levels = change_levels)
  )

# Fill scale
fill_scale <- scale_fill_distiller(
  palette = "RdYlGn",
  direction = 1,
  na.value = "black"
)

# Base theme
base_theme <- theme_minimal(base_family = "Times New Roman") +
  theme(
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    strip.text = element_text(color = "black", size = 13)
  )


# Histogram of ChangeClass faceted by model
rs_df_2 <- raster_df_processed_clean%>% filter(Model == "RS")
landis_df_2 <- raster_df_processed_clean%>% filter(Model == "LANDIS")
dual_plots_rs <- list()
dual_pct_plots_rs <- list()
dual_plots_landis <- list()
dual_pct_plots_landis <- list()
combined_plots_all <- list()
for (spp in species){
  
  rs_plot <- ggplot() +
    geom_raster(data = rs_df_2%>%filter(Species==c("Alaskan birch", "Trembling aspen")), aes(x = x, y = y, fill = Value)) +
    geom_raster(data = zero_df, aes(x = x, y = y), fill = "black") +
    geom_sf(data = bounds_sf, fill = NA, color = "black", linewidth = 0.3) +
    fill_scale +
    coord_sf(expand = FALSE) +
    #facet_grid(Species~.) +
    #labs(fill = "Δ C\n(g m-2)")+
    base_theme +
    theme(legend.position = "none",
          strip.text = element_blank())  
  #rs_plots[[spp]] <- rs_plot

  hist_plot <- ggplot(raster_df_processed_clean%>%filter(Species==spp, Model=="RS"), 
                      aes(x = ChangeClass, fill = ChangeClass)) +
    geom_bar() +
    #facet_wrap(~Model) +
    scale_fill_brewer(palette = "RdYlGn", na.value = "black") +
    labs(
      #x = "Change in Carbon (g C / m²)",
      y = "Pixel Count",
      fill = "ΔC"
    ) +
    coord_flip()+
    theme_minimal(base_family = "Times New Roman") +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      strip.text = element_text(color = "black", size = 13),
      legend.position = "none"
    )
  
  pct_base_rs <- ggplot(change_summary %>% filter(Species == spp),
                     aes(x = ChangeClass, y = percent, fill = ChangeClass)) +
    geom_bar(stat = "identity") +
    facet_grid(Species ~ .) +
    scale_fill_brewer(palette = "RdYlGn", na.value = "black") +
    coord_cartesian(xlim = c("< -100", "> 100"), ylim = c(0, 40))+
    coord_flip() +
    theme_minimal(base_family = "Times New Roman") +
    theme(
      axis.text.x = element_text(size = 16, hjust = 1),
      axis.text.y = element_text(size = 16, hjust = 1),
      axis.title = element_text(size = 18, color ="black"),
      strip.text = element_text(color = "black", size = 18),
      legend.position = "none"
    )
  
  # Add labels conditionally
  if (spp == spp) {
    pct_plot <- pct_base_rs +
      labs(y = "Percent area", x = "ΔC (g C m-2)")
  } else {
    pct_plot <- pct_base_rs +
      labs(y=element_blank(),x = "ΔC (g C m-2)")
  }

  dual_plot <- rs_plot | hist_plot+plot_layout(widths = c(1, .5))
  dual_plots_rs[[spp]] <- dual_plot
  
  dual_plot_pct <- rs_plot | pct_plot +plot_layout(widths = c(1, .25))
  dual_pct_plots_rs[[spp]] <- dual_plot_pct
  
  
  #landis
    
  landis_plot <- ggplot() +
    geom_raster(data = landis_df_2%>%filter(Species==spp), aes(x = x, y = y, fill = Value)) +
    geom_raster(data = zero_df, aes(x = x, y = y), fill = "black") +
    geom_sf(data = bounds_sf, fill = NA, color = "black", linewidth = 0.3) +
    fill_scale +
    coord_sf(expand = FALSE) +
    facet_grid(Species~.) +
    #labs(fill = "Δ C\n(g m-2)")+
    base_theme +
    theme(legend.position = "none",
          strip.text = element_blank())  
  #landis_plots[[spp]] <- landis_plot

 
  
  pct_base <- ggplot(change_summary_2 %>% filter(Species == spp),
                     aes(x = ChangeClass, y = percent, fill = ChangeClass)) +
    geom_bar(stat = "identity") +
    facet_grid(Species ~ .) +
    scale_fill_brewer(palette = "RdYlGn", na.value = "black") +
    coord_cartesian(xlim = c("< -100", "> 100"), ylim = c(0, 50))+
    coord_flip() +
    theme_minimal(base_family = "Times New Roman") +
    theme(
      axis.text.x = element_text(size = 16, hjust = 1),
      axis.text.y = element_text(size = 16, hjust = 1),
      axis.title = element_text(size = 18, color ="black"),
      strip.text = element_text(color = "black", size = 18),
      legend.position = "none"
    )
  
  
  # Add labels conditionally
  if (spp == spp) {
    pct_plot_landis <- pct_base +
      labs(y = "Percent area", x = "ΔC (g C m-2)")
  } else {
    pct_plot_landis <- pct_base +
      labs(y= element_blank(),x = "ΔC (g C m-2)")
  }
  
  dual_plot_pct_landis<- landis_plot | pct_plot_landis +plot_layout(widths = c(1, .25))
  dual_pct_plots_landis[[spp]] <- dual_plot_pct_landis
  full_combined <- rs_plot | pct_plot + plot_layout(widths = c(4, .5))| landis_plot | pct_plot_landis + plot_layout(widths = c(4, .5))
  combined_plots_all[[spp]] <- full_combined
}
# Final output
dual_plots_rs_reduced <- reduce(dual_plots_rs, `/`)
dual_pct_plots_rs_reduced <- reduce(dual_pct_plots_rs, `/`)
#dual_plots_landis_reduced <- reduce(dual_plots_landis, `/`)
dual_pct_plots_landis_reduced <- reduce(dual_pct_plots_landis, `/`)
combined_plots_reduced <- reduce(combined_plots_all, `/`)

```

```{r}
png("figures/Results_MAPAS_2.png", width = 11, height = 10, units = "in", res = 300)
print(combined_plots_reduced)
dev.off()
```

relationships between positive increases between models

```{r}
library(dplyr)
library(tidyr)
library(purrr)
library(broom)

# Add a unique pixel ID per species-model combo (assumes row order implies spatial alignment)
raster_df_processed_clean <- raster_df_processed_clean %>%
  group_by(Model, Species) %>%
  mutate(RowID = row_number()) %>%
  ungroup()

# Reshape data so each row is a pixel with RS and LANDIS values side-by-side
wide_df <- raster_df_processed_clean %>%
  select(Species, Model, RowID, Value) %>%
  pivot_wider(names_from = Model, values_from = Value) %>%
  filter(!is.na(RS) & !is.na(LANDIS)) %>%
  mutate(rs_change = RS > 0,
         landis_change = LANDIS > 0)

# Run logistic regression (or linear) for each species
model_results <- wide_df %>%
  group_by(Species) %>%
  nest() %>%
  mutate(
    model = map(data, ~ lm(rs_change ~ landis_change, data = .x)),
    summary = map(model, glance),
    coefficients = map(model, tidy)
  ) %>%
  select(Species, summary, coefficients) %>%
  unnest(summary)

# View results
model_results %>%
  select(Species, r.squared, p.value, adj.r.squared, statistic)
species_list <- unique(raster_df_processed_clean$Species)

agreement_summary <- lapply(species_list, function(species) {
  rs_vals <- raster_df_processed_clean$Value[raster_df_processed_clean$Species == species & 
                                             raster_df_processed_clean$Model == "RS"]
  landis_vals <- raster_df_processed_clean$Value[raster_df_processed_clean$Species == species & 
                                                 raster_df_processed_clean$Model == "LANDIS"]
  
  # Must ensure length match
  n <- min(length(rs_vals), length(landis_vals))
  rs_gain <- rs_vals[1:n] > 0
  landis_gain <- landis_vals[1:n] > 0
  
  # Agreement vectors
  agreement <- rs_gain == landis_gain
  
  list(
    Species = species,
    AgreementPercent = mean(agreement) * 100,
    GainBoth = mean(rs_gain & landis_gain) * 100,
    LossBoth = mean(!rs_gain & !landis_gain) * 100,
    RS_Gain_LANDIS_Loss = mean(rs_gain & !landis_gain) * 100,
    RS_Loss_LANDIS_Gain = mean(!rs_gain & landis_gain) * 100
  )
})
agreement_df <- do.call(rbind, lapply(agreement_summary, as.data.frame))

##does conifer loss correspond to hardwood gain (models independtly)
library(tidyr)
conifers <- c("Black spruce", "White spruce")
hardwoods <- c("Alaskan birch", "Trembling aspen")
wide_df <- raster_df_processed_clean %>%
  pivot_wider(names_from = c(Model, Species), values_from = Value)
rs_conifer_loss <- (wide_df$`RS_Black spruce` < 0) | (wide_df$`RS_White spruce` < 0)

# Hardwood gains in RS
rs_hardwood_gain <- (wide_df$`RS_Alaskan birch` > 0) | (wide_df$`RS_Trembling aspen` > 0)

# % of conifer-loss pixels where hardwoods increased
rs_overlap <- mean(rs_conifer_loss & rs_hardwood_gain, na.rm = TRUE) * 100
landis_conifer_loss <- (wide_df$`LANDIS_Black spruce` < 0) | (wide_df$`LANDIS_White spruce` < 0)
landis_hardwood_gain <- (wide_df$`LANDIS_Alaskan birch` > 0) | (wide_df$`LANDIS_Trembling aspen` > 0)

landis_overlap <- mean(landis_conifer_loss & landis_hardwood_gain, na.rm = TRUE) * 100
rs_conifer_loss_percent <- mean(rs_conifer_loss, na.rm = TRUE) * 100
landis_conifer_loss_percent <- mean(landis_conifer_loss, na.rm = TRUE) * 100
rs_hardwood_gain_percent <- mean(rs_hardwood_gain, na.rm = TRUE) * 100
landis_hardwood_gain_percent <- mean(landis_hardwood_gain, na.rm = TRUE) * 100
data.frame(
  Metric = c("Conifer Loss", "Hardwood Gain", "Conifer Loss + Hardwood Gain"),
  `RS-RF (%)` = c(round(rs_conifer_loss_percent, 1),
                  round(rs_hardwood_gain_percent, 1),
                  round(rs_overlap, 1)),
  `LANDIS-II (%)` = c(round(landis_conifer_loss_percent, 1),
                      round(landis_hardwood_gain_percent, 1),
                      round(landis_overlap, 1))
)
```
