---
title: "Random Forest Modeling Landis Biomass"
author: "Wesley Rancher"
date: "2025-04-28"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
dir <- "F:/randomForest_Landis/"
dir <- "/Users/wancher/Documents/thesis/data/analysis/"
setwd(dir)
```

Read in training dataset and rasters to make predictions on

```{r}
library(dplyr)
#Training Dataframe was prepared in (../scripts/DataPreparation.Rmd)
#training_df <- read.csv("data/analysis/main_predictor_data_05272025.csv") 
training_df <- read.csv("main_predictor_data_05272025.csv")
```

Define some functions to make predictions and plot these estimates against our validation set. I'm using gsub and  strsplit to grab unique identifiers from file names so I can store them in a dataframe for later plotting. This requires a final fit from a model workflow. These functions are nested in the model training function.


```{r}
make_validation_df <- function(fit, test_data, species, mean_rmse, mean_rsq){
  spp <- tools::toTitleCase(strsplit(species, " ")[[1]][1])%>%paste0(" ", strsplit(species, " ")[[1]][2])
  predicted <- predict(fit, test_data)
  observed <- test_data$Carbon_gm2
  vals <- predicted[[1]]
  df <- rbind(data.frame(Observed=observed, Predicted=vals, Species=spp, RMSE=mean_rmse, RSQ=mean_rsq))
  return(df)
}
```

Function to conduct multidimensional scaling on the proximity matrix generated from the random forests.

```{r}
library(vegan)
nmds_rf <- function(raw_model, filtered_train_df, species) {
  if (is.null(raw_model$proximity)) {
    stop("Random Forest model must be trained with proximity=TRUE to use MDS.")
  }
  
  prox <- raw_model$proximity
  
  #nmds
  mds_coords <- monoMDS(1 - prox, k = 2, maxit=100, model="local")
  mds_df <- as.data.frame(scores(mds_coords, display  = "sites"))
  #colnames(mds_df) <- c("Dim1", "Dim2")
  
  mds_df$Species <- species
  mds_df$ag_carbon <- filtered_train_df$ag_carbon
  mds_df$stress <- mds_coords$stress#for reporting
  
  #join to predictors
  numeric_predictors <- filtered_train_df %>% 
    select(where(is.numeric), -spp, -PSP, -scenario, -AvailableWater)
  
  #environmental fit
  ef <- envfit(mds_coords, numeric_predictors, permutations = 999)
  ef_df <- as.data.frame(scores(ef, "vectors"))
  ef_df$Feature <- rownames(ef_df)
  ef_df$pval <- ef$vectors$pvals
  ef_df$Species <- species

  return(list(mds=mds_df, envfit=ef_df))
}
```

Model fitting 

```{r}
library(purrr)
library(tidymodels)
library(randomForest)
#library(partykit)
library(vip)
library(rlang)

species <- c("BlackSpruce", "WhiteSpruce", "PaperBirch", "QuakingAspen")

train_model <- function(df, species, response="ag_carbon") {
  print(paste0("running model for: ", species))
  
  #filter to one species
  training_df_one_spp <- training_df %>% filter(spp == species)
  predictors <- setdiff(names(training_df_one_spp), c("spp", "scenario", "PSP", response))

  #train and validation set (stratify the predictor dataset based on nature of the data measured in situ)
  split_data <- initial_split(training_df_one_spp %>% select(all_of(response), all_of(predictors)), prop = 0.95, strata = all_of(response))
  
  #training and test set
  train_data <- training(split_data)
  test_data <- testing(split_data)
  formula <- as.formula(paste(response, "~", paste(predictors, collapse = "+")))
  
  #tidy models recipe (swapping in entire dataset)
  recipe <- recipe(formula = formula, data = training_df_one_spp) %>% 
    #step_nzv(all_predictors()) %>%
    step_corr(all_numeric_predictors(), threshold = 0.80, use = "pairwise.complete.obs", method = "pearson")
  
  #define models to run (nts: cite packages)
  #use additional script to set up cforest to be tidymodels friendly
  #source("scripts/define_cforest_model.R")
  m_rf <- rand_forest(mtry = tune(), trees = tune(), min_n = tune()) %>% set_mode("regression") %>% set_engine("randomForest", proximity=TRUE)
  m_ranger <- rand_forest(mtry = tune(), trees = tune(), min_n = tune()) %>% set_mode("regression") %>% set_engine("ranger")
  m_xgb <- boost_tree(mtry = tune(), trees = tune(), min_n = tune()) %>% set_mode("regression") %>% set_engine("xgboost")
  m_kknn <- nearest_neighbor(neighbors=tune(), weight_func = tune(), dist_power = tune()) %>% set_mode("regression") %>% set_engine("kknn")
  m_svm <- svm_rbf(cost= tune(), rbf_sigma = tune()) %>% set_mode("regression") %>% set_engine("kernlab")
  #m_cforest <- cforest() %>% set_mode("regression") %>% set_engine("partykit")
  models <- list(rf=m_rf,ranger=m_ranger)
  
  #params and grids
  #top3_params <- parameters(mtry(), trees(), min_n())
  # I think these two need recipes where the data is normalized
  kknn_params <- parameters(neighbors(range = c(4, 25)), weight_func(values = c("rectangular", "triangular", "biweight", 
                                                                                "triweight", "epanechnikov", "gaussian")),dist_power(range = c(1, 2)))  
  svm_params <- parameters(cost(range = c(-8, 4), trans = scales::log10_trans()), rbf_sigma(range = c(-3, 0), trans = scales::log10_trans()))
    
  #rf, ranger, and xgb use the same params
  rf_grid <- expand.grid(mtry = seq(5, 20, 5), trees = seq(100,800,200), min_n = seq(1, 12, 4))

  kknn_grid <- grid_regular(kknn_params, levels = c(4,3,2))
  svm_grid <- grid_regular(svm_params, levels = 10)
  grids <- list(rf_grid=rf_grid, ranger_grid=rf_grid)
  
  #cross validation
  nfold <- 3
  print(paste0("performing cross validation across ", nfold, " folds"))
  pv_folds <- vfold_cv(training_df_one_spp, v = nfold, strata = all_of(response))

  #store the results for all models (selecting the best model across the cross-val folds based on RMSE)
  #working outside of the map2 function to generate predictions
  
  #results <- map2(models, grids, function(model, grid) {
  #trying to leave this dynamic
  tuned_results <- workflow() %>% add_recipe(recipe) %>% add_model(m_rf) %>% 
    tune_grid(resamples = pv_folds, grid = rf_grid, metrics = metric_set(rmse, rsq, mae))
  autoplot(tuned_results)
  best_params <- select_best(tuned_results, metric="rmse")
  wflow <- finalize_workflow(workflow() %>% add_recipe(recipe) %>% add_model(m_rf), best_params)
  
  #fit across the resamples
  fit <- fit_resamples(wflow, resamples = pv_folds, metrics = metric_set(rmse, rsq, mae))
  mean_rmse <- collect_metrics(fit) %>% filter(.metric == "rmse") %>% pull(mean)
  mean_rsq <- collect_metrics(fit) %>% filter(.metric == "rsq") %>% pull(mean)
  
  #print(paste0("MODEL: ", model$engine))
  print(paste("Mean RMSE across folds: ", mean_rmse))
  print(paste("Mean R² across folds: ", mean_rsq))
  
  final_fit <- fit(wflow, training_df_one_spp)
  
  #nmds
  raw_model <- extract_fit_engine(final_fit)
  mds_results <- nmds_rf(raw_model, training_df_one_spp, species)
  
  #plot for one species
  # mdsplot <- ggplot(mds_df, aes(x = MDS1, y = MDS2, color = ag_carbon)) +
  #   geom_point(size=4,alpha=0.8) +
  #   scale_color_viridis_c(option="turbo",direction=1)+
  #   theme_bw()+
  #   labs(title=paste0(species))+
  #   theme(legend.position="bottom")
  # print(mdsplot)
  
  #extract
  finalized_model <- extract_fit_parsnip(final_fit)
  
  ntree <-  quo_get_expr(finalized_model[[2]]$args$trees)
  mtry <- quo_get_expr(finalized_model[[2]]$args$mtry)
  min_number <- quo_get_expr(finalized_model[[2]]$args$min_n)
  
  print(paste("Number of trees: ", ntree))
  print(paste("mtry: ", mtry))
  print(paste("min n", min_number))
  
  # importance scores (permuation based)
  importance <- vi(finalized_model)
  importance$Species <- species
  #stored_plt <- vip(importance, num_features = 10, geom = "point")
  
  #  return(final_fit)
  #})
  
  #generate predictions (using the regular RF model fit)
  #all_predictions <- lapply(unique(r_files), function(file) {predict_from_model(file, final_fit, species)})
  #all_predictions_df <- bind_rows(all_predictions)
  
  #generate validation data for plotting
  #validation_fit <- fit(wflow, test_data)
  #validation_pred_df <- make_validation_df(validation_fit, test_data, species, mean_rmse, mean_rsq)
    
  return(list(train_dfs=NULL, 
              test_dfs=NULL,
              mds_data=mds_results,
              final_models=finalized_model,
              varImp_data=importance, 
              all_predictions=NULL, 
              validation_data=NULL))
  #rm(all_predictions, all_predictions_df)
}

#run the models
model_output <- lapply(species, function(spp){train_model(training_df,species=spp)})
varImp_all_spp <- do.call(rbind, lapply(model_output[1:4], function(x) x$varImp_data))
#predictions <- do.call(rbind, lapply(model_output[1:4], function(x) x$all_predictions))
#validations <- do.call(rbind, lapply(model_output[1:4], function(x) x$validation_data))
#metrics <- do.call(rbind, lapply(model_output[1:4], function(x) x$final_models))
proximity_data <- do.call(rbind, lapply(model_output[1:4], function(x) x$mds_data[[1]]))
environmental_fit_data <- do.call(rbind, lapply(model_output[1:4], function(x) x$mds_data[[2]]))

#write.csv(predictions, "F:/jetstream/data/PredictedCarbon_RasterValues_04252025.csv", row.names = FALSE)
#write.csv(varImp_all_spp, "F:/jetstream/data/VariableImportance_LANDIS_04282025.csv", row.names = FALSE)
#write.csv(proximity_data, "data/analysis/MDS_Data_LANDIS_04292025.csv", row.names = FALSE)
#write.csv(environmental_fit_data, "data/analysis/EnvFit_Data_LANDIS_04292025.csv", row.names = FALSE)

```

Graph the predicted vs observed data for the validation fits.

```{r}
library(ggplot2)
library(viridis)

#plotting validation data vs predictions for each species with a 60/40 split
validation_plot <- ggplot(validations, aes(x = Observed, y = Predicted, color = abs(Predicted-Observed))) +
  #geom_smooth(method = "lm", color = "black", linetype = "dashed", se = FALSE)+
  geom_point(size = 5, alpha = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "solid") +
  scale_color_viridis(option = "H", direction=-1)+
  facet_grid(~ Species, nrow(1)) +
  theme_bw() +
  scale_x_continuous(limits=c(0,500))+
  scale_y_continuous(limits=c(0,500))+
  labs(x = "Observed", y = "Predicted") +
  theme(legend.position = "none", strip.text = element_text(size = 14), axis.title = element_text(size = 16)) +
  geom_text(aes(x = 0.9 * max(Observed), y = 0.1 * max(Predicted), label = paste("r2:", round(RSQ, 2), "\nRMSE:", round(RMSE, 2))),
            size = 10, hjust = 1, vjust = 0, color = "black")

plot(validation_plot)
```

Graph variable importance

```{r}
#library(forcats)
colors <- c("Alaskan birch" = "#67161CFF","Black spruce" = "#3F6148FF",
            "White spruce" = "#A4804CFF","Trembling aspen" = "#4B5F80FF")

varImp_all_spp%>%
  
  mutate(Species = ifelse(Species == "PaperBirch", "Alaskan birch", Species),
         Species = ifelse(Species == "QuakingAspen", "Trembling aspen", Species),
         Species = ifelse(Species == "BlackSpruce", "Black spruce", Species),
         Species = ifelse(Species == "WhiteSpruce", "White spruce", Species))%>%
  #filter(Species=="Trembling aspen")%>%
  group_by(Species) %>%
  slice_max(order_by = Importance, n = 7, with_ties = FALSE) %>%
  #arrange(Species, desc(Importance))%>%
  ungroup() %>%
  ggplot(aes(x=Importance, y=Variable, fill=Species))+
  #geom_point()+
  geom_bar(stat="identity")+
  #geom_line(linewidth=1)+
  scale_color_manual(values=colors)+
  scale_fill_manual(values = colors)+
  #scale_x_continuous(limits=c(2000,2025), breaks = 5)+
  facet_grid(Species~., scales = "free_y", space = "free_y")+
  theme_bw()+
  theme(legend.position="none")
```
